Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2005-04-10 paper/metamonads.tex
corrections to the code

In addition to the ``regular'' type sharing constraints shown in the
|Gen| functor, there are also ``semantic'' sharing constraints, shown
in the following structure of the |UPDATE| signature:
\vspace*{-5pt}\begin{code}
module DivisionUpdate
  (Dom:DOMAIN with type kind = domain_is_field)
  (C:CONTAINER2D)
  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
\end{code}
\vspace*{-4pt} 
This structure implements an update policy of using
|Dom.div| operation without restrictions -- which is possible only if
the domain has such unrestricted operation. A domain such as the integer domain
may still provide |Dom.div| of the same type, but that operation may
only be used when we are sure that the division is exact. Our type
sharing constraint expresses such domain-specific knowledge:
instantiating |DivisionUpdate| with |integerDomain| leads to a
compile-time error, when compiling the \emph{generator} code. Thus, in
some cases we can use module types for ``semantic'' constraints
that cannot normally be expressed via the types of module members.
\vspace*{-5pt}
\begin{code}
module GenFV4 = Gen(FloatDomain)
   (GenericVectorContainer)(FullPivot)
   (DivisionUpdate(FloatDomain)(GenericVectorContainer)(FDet))
   (OutDetRank(FloatDomain)(GenericVectorContainer)(FDet)(Rank))
module GenIA1 = Gen(IntegerDomain)
   (GenericArrayContainer)(RowPivot)
   (FractionFreeUpdate(IntegerDomain)(GenericArrayContainer)(IDet))
   (OutJustMatrix(IntegerDomain)(GenericArrayContainer)(IDet))
\end{code}
\vspace*{-5pt}
We can instantiate the |Gen| functor as shown above and inspect the
generated code, e.g., by printing |GenIA1.gen|. The code can then be
``compiled'' as |!. GenIA1.gen| or with off-shoring. The code
% (cite).  No space!
for |GenFV4| shows full pivoting, determinant and rank tracking. The
code for all these aspects is fully inlined; no extra functions are
invoked and no tests other than those needed by the GE algorithm
itself are performed. The GE function returns a triple 
|float array array * float * int| of the U-factor, determinant and 
the rank. The
code generated by |GenIA1| shows absolutely no traces of determinant
tracking: no declaratrion of spurious variables, no extra tests,
etc. The code appears as if the determinant tracking aspect did not exist
at all. The generated code for the above and other instantiations of
|Gen| can be examined at \cite{metamonadsURL}.

@@

\bibliography{metamonads}
\bibliographystyle{plain}
\end{document}

% It seems no appendices are allowed. Check with Walid?

>>>>

In addition to the ``regular'' type sharing constraints shown in the
|Gen| functor, there are also ``semantic'' sharing constraints, shown
in the following structure of the |UPDATE| signature:
\vspace*{-5pt}\begin{code}
module DivisionUpdate
  (Dom:DOMAIN with type kind = domain_is_field)
  (C:CONTAINER2D)
  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
\end{code}
\vspace*{-4pt} 
This structure implements an update policy of using
|Dom.div| operation without restrictions -- which is possible only if
the domain has such unrestricted operation. A domain such as the integer domain
may still provide |Dom.div| of the same type, but that operation may
only be used when we are sure that the division is exact. Our type
sharing constraint expresses such domain-specific knowledge:
instantiating |DivisionUpdate| with |integerDomain| leads to a
compile-time error, when compiling the \emph{generator} code. Thus, in
some cases we can use module types for ``semantic'' constraints
that cannot normally be expressed via the types of module members.
\vspace*{-5pt}
\begin{code}
module GenIV5 = Gen(IntegerDomain)
   (GenericVectorContainer)(FullPivot)
   (FractionFreeUpdate(IntegerDomain)(GenericVectorContainer)(IDet))
   (OutDetRank(IntegerDomain)(GenericVectorContainer)(IDet)(Rank))
module GenFA1 = Gen(FloatDomain)
   (GenericArrayContainer)(RowPivot)
   (DivisionUpdate(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
   (OutJustMatrix(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
\end{code}
\vspace*{-5pt}
We can instantiate the |Gen| functor as shown above and inspect the
generated code, e.g., by printing |GenFA1.gen|. The code can then be
``compiled'' as |!. GenFA1.gen| or with off-shoring. The code
for |GenIV5| (Appendix A) shows full pivoting, determinant and rank tracking. The
code for all these aspects is fully inlined; no extra functions are
invoked and no tests other than those needed by the GE algorithm
itself are performed. The GE function returns a triple 
|int array * int * int| of the U-factor, determinant and 
the rank. The
code generated by |GenFA1| (Appendix B) shows absolutely no traces of determinant
tracking: no declaration of spurious variables, no extra tests,
etc. The code appears as if the determinant tracking aspect did not exist
at all. The generated code for the above and other instantiations of
|Gen| can be examined at \cite{metamonadsURL}.

@@

\bibliography{metamonads}
\bibliographystyle{plain}
\section{Appendix A}
The code generated for |GenIV5|, fraction-free GE of the integer matrix
represented by a flat vector, full pivoting, returning the |U|-factor,
the determinant and the rank.
\begin{code}
# val resIV5 :
  ('a,
   Funct4.GenIV5.Ctr.contr ->
   Funct4.OutDetRank(Funct4.IntegerDomain)(Funct4.GenericVectorContainer)(Funct4.IDet)(Funct4.Rank).res)
  code =
  .<fun a_405 ->
   let t_406 = (ref 0) in
   let t_407 = (ref 0) in
   let t_408 = {arr = (Array.copy a_405.arr)} (a_405) in
   let t_409 = a_405.m in
   let t_410 = a_405.n in
   let t_411 = (ref 1) in
   let t_412 = (ref 1) in
   while (((! t_407) < t_409) && ((! t_406) < t_410)) do
    let t_413 = (! t_406) in
    let t_414 = (! t_407) in
    let t_415 = (ref (None)) in
    let t_435 =
     begin
      for j_431 = t_413 to (t_410 - 1) do
       for j_432 = t_414 to (t_409 - 1) do
        let t_433 = (t_408.arr).((j_431 * t_408.m) + j_432) in
        if (not (t_433 = 0)) then
         (match (! t_415) with
          | Some (i_434) ->
             if ((abs (snd i_434)) > (abs t_433)) then
              (t_415 := (Some ((j_431, j_432), t_433)))
             else ()
          | None -> (t_415 := (Some ((j_431, j_432), t_433))))
        else ()
       done
      done;
      (match (! t_415) with
       | Some (i_416) ->
          if ((snd (fst i_416)) <> t_414) then begin
           let a_424 = t_408.arr
           and nm_425 = (t_408.n * t_408.m)
           and m_426 = t_408.m in
           let rec loop_427 =
            fun i1_428 ->
             fun i2_429 ->
              if (i2_429 < nm_425) then
               let t_430 = a_424.(i1_428) in
               a_424.(i1_428) <- a_424.(i2_429);
               a_424.(i2_429) <- t_430;
               (loop_427 (i1_428 + m_426) (i2_429 + m_426))
              else () in
           (loop_427 t_414 (snd (fst i_416)));
           (t_412 := (~- (! t_412)))
          end else ();
          if ((fst (fst i_416)) <> t_413) then begin
           let a_417 = t_408.arr
           and n_418 = t_408.n
           and m_419 = t_408.m in
           let i1_420 = (t_413 * m_419)
           and i2_421 = ((fst (fst i_416)) * m_419) in
           for i_422 = 0 to (m_419 - 1) do
            let t_423 = a_417.(i1_420 + i_422) in
            a_417.(i1_420 + i_422) <- a_417.(i2_421 + i_422);
            a_417.(i2_421 + i_422) <- t_423
           done;
           (t_412 := (~- (! t_412)))
          end else ();
          (Some (snd i_416))
       | None -> (None))
     end in
    (match t_435 with
     | Some (i_436) ->
        begin
         for j_437 = (t_413 + 1) to (t_410 - 1) do
          if (not ((t_408.arr).((j_437 * t_408.m) + t_414) = 0)) then begin
           for j_438 = (t_414 + 1) to (t_409 - 1) do
            (t_408.arr).((j_437 * t_408.m) + j_438) <-
             ((((t_408.arr).((j_437 * t_408.m) + j_438) *
                 (t_408.arr).((t_413 * t_408.m) + t_414)) -
                ((t_408.arr).((t_413 * t_408.m) + j_438) *
                  (t_408.arr).((j_437 * t_408.m) + t_413))) / (! t_411))
           done;
           (t_408.arr).((j_437 * t_408.m) + t_414) <- 0
          end else ()
         done;
         (t_411 := i_436)
        end;
        (t_406 := ((! t_406) + 1))
     | None -> (t_412 := 0));
    (t_407 := ((! t_407) + 1))
   done;
   (t_408,
    if ((! t_412) = 0) then 0
    else if ((! t_412) = 1) then (! t_411)
    else (~- (! t_411)), (! t_406))>.
\end{code}
\section{Appendix B}
The code generated for |GenFA1|, GE of the floating point matrix
represented by a 2D array, row pivoting, returning just the |U|-factor.
\begin{code}
#   val resFA1 :
  ('a,
   Funct4.GenFA1.Ctr.contr ->
   Funct4.OutJustMatrix(Funct4.FloatDomain)(Funct4.GenericArrayContainer)(Funct4.NoDet(Funct4.FloatDomain)).res)
  code =
  .<fun a_1 ->
   let t_2 = (ref 0) in
   let t_3 = (ref 0) in
   let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
   let t_6 = (Array.length a_1.(0)) in
   let t_7 = (Array.length a_1) in
   while (((! t_3) < t_6) && ((! t_2) < t_7)) do
    let t_8 = (! t_2) in
    let t_9 = (! t_3) in
    let t_10 = (ref (None)) in
    let t_16 =
     begin
      for j_13 = t_8 to (t_7 - 1) do
       let t_14 = (t_5.(j_13)).(t_9) in
       if (not (t_14 = 0.)) then
        (match (! t_10) with
         | Some (i_15) ->
            if ((abs_float (snd i_15)) < (abs_float t_14)) then
             (t_10 := (Some (j_13, t_14)))
            else ()
         | None -> (t_10 := (Some (j_13, t_14))))
       else ()
      done;
      (match (! t_10) with
       | Some (i_11) ->
          if ((fst i_11) <> t_8) then begin
           let t_12 = t_5.(t_8) in
           t_5.(t_8) <- t_5.(fst i_11);
           t_5.(fst i_11) <- t_12;
           ()
          end else ();
          (Some (snd i_11))
       | None -> (None))
     end in
    (match t_16 with
     | Some (i_17) ->
        begin
         for j_18 = (t_8 + 1) to (t_7 - 1) do
          if (not ((t_5.(j_18)).(t_9) = 0.)) then begin
           for j_19 = (t_9 + 1) to (t_6 - 1) do
            (t_5.(j_18)).(j_19) <-
             ((t_5.(j_18)).(j_19) -.
               (((t_5.(j_18)).(t_9) /. (t_5.(t_8)).(t_9)) *.
                 (t_5.(t_8)).(j_19)))
           done;
           (t_5.(j_18)).(t_9) <- 0.
          end else ()
         done;
         ()
        end;
        (t_2 := ((! t_2) + 1))
     | None -> ());
    (t_3 := ((! t_3) + 1))
   done;
   t_5>.
\end{code}
\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% It seems no appendices are allowed. Check with Walid?
